\vspace{1.0cm} \Large \bf Supplemental material of ``Health and disease imprinted in the time variability of the human microbiome''

### 2016-07-26 mSystems/mSys_discussion.tex
Several minor changes leading to 0.5.0, another step towards pre-releasing
On branch master changes to be committed:
	modified:   mSys_MAIN.pdf
	modified:   mSys_MAIN.tex
	modified:   mSys_MandM.tex
	modified:   mSys_discussion.tex
	modified:   mSys_intro.tex
	modified:   mSys_results.tex

The primary hypothesis of this work is that, in adults, having a healthy microbiota means that population is stable in time and does not have huge flips or jumps into another states. In order to use the valuable information which gives us the empirical law of Taylor's work, we propose the use of Langevin equation to model how the ranking stability evolves in time. While we can measure directly the component of the noise of the system as their variability, the other main term needs to be inferred from the model. This term, which we have named as 'fitness', is the one that gives the ability to the system to be stable to potential perturbations. In ecological terms, this could mean the nature of interactions that are present among the bacteria, between bacteria and other minority populations as fungi or archaea, between bacteria and the viral component in the microbiota, and the interactions between host and the whole microbiota. Being this a first step to model the temporal stability of the microbiota and due to its complicated nature, we have calculated the fitness term using the fluctuation-dissipation theorem as a first approximation. Thus, the fitness of the microbiota still remains to be modelled in future works in order to make the model more accurate and with a higher predictive power.

>>>>

The primary hypothesis of this work is that, in adults, having a healthy microbiota means that population is stable in time and does not have huge flips or jumps into another states. In order to use the valuable information which gives us the empirical law of Taylor's work, we propose the use of Langevin equation to model how the ranking stability evolves in time. While we can measure directly the component of the noise of the system as their variability, the other main term needs to be inferred from the model. This term, which we have named as 'fitness', is the one that gives the ability to the system to be stable to potential perturbations. In ecological terms, this could mean the nature of interactions that are present among the bacteria, between bacteria and other minority populations as fungi or archaea, between bacteria and the viral component in the microbiota, and the interactions between host and the whole microbiota. Being this a first step to model the temporal stability of the microbiota and due to its complicated nature, we have calculated the fitness term using the Fluctuation Dissipation Theorem as a first approximation\cite{FD}. Thus, the fitness of the microbiota still remains to be modelled in future works in order to make the model more accurate and with a higher predictive power.

### 2016-07-26 mSystems/mSys_results.tex
Several minor changes leading to 0.5.0, another step towards pre-releasing
On branch master changes to be committed:
	modified:   mSys_MAIN.pdf
	modified:   mSys_MAIN.tex
	modified:   mSys_MandM.tex
	modified:   mSys_discussion.tex
	modified:   mSys_intro.tex
	modified:   mSys_results.tex

%%%
\subsection*{Rank stability of the taxa} 
The rank dynamics and stability plot in Figure \ref{fig:corrank} shows the variation in the rank with time for the most dominant taxa and their calculated Rank Stability Index (RSI, as discussed in Material and Methods) for the taxa of a healthy subject (individual \emph{A}, top) and from a subject diagnosed with IBS (patient \emph{P2}, bottom) of the IBS study\cite{IBS}. The taxa are listed ordered by the accumulated frequency along the time series, so y-axis is an overall dominance axis for each sample set. Generally speaking, we observe that the most dominant taxa are the most rank stable.

\begin{figure}
	\centering
	\vspace*{-10mm} % Corrects overbox of the figures
	\includegraphics[width=0.8\textwidth]{results/corrank/IBS_h_A_amplicons_family_Rank.pdf}\vspace*{-3mm} % Corrects overbox of the figures
	\hspace*{-4.5mm}\includegraphics[width=0.83\textwidth]{results/corrank/IBS_P2_Metatranscriptores_Rank.pdf}
	\caption{Matrixes showing the rank variation throughout time for the most dominant elements (taxa) and their calculated Rank Stability Index (as shown in Material and Methods). We show the matrix for samples from a healthy subject (top) and from a subject diagnosed with irritable bowel syndrome (bottom), studied in our lab\cite{IBS}.}
	\label{fig:corrank}
\end{figure}

Nevertheless, in the particular case of the healthy individual, \emph{Burkholderiales} and \emph{Betaproteobacteria} (taxa ordered as 18th and 25th in the dominance axis) show comparatively very low rank stability regarding similar dominant taxa while, on the other hand, \emph{Comamonadaceae}, \emph{Lactobacillaceae}, \emph{Fusobacteriaceae}, \emph{Aerococcaceae} and \emph{Carnobacteriaceae} show higher stability than other more dominant taxa, forming a kind of \emph{rank stability island} for medium-ranked taxa around position 40 in the dominance axis, and thus colored in orange following Table \ref{tab:RSI} criteria, since they show a moderately stable RSI.

\begin{table}
  \begin{center}
    \begin{tabular}{cccr}
    \hline
    Case  &  Condition  &  Colour  &  Description \\
    \hline
    1  &  $1\ge{\rm RSI}>0.99$  & \textcolor{blue}{blue} & constant rank \\ 
    2  & ${\rm RSI}>0.90$  &  \textcolor{green}{green}  & highly stable rank \\
    3  &  ${\rm RSI}>0.75$  &  \textcolor{orange}{orange} & moderately stable rank \\
    4  &  ${\rm RSI}>0.25$  &  \textcolor{red}{red} & unstable rank \\
    5  &  $0.25\ge{\rm RSI}\ge0$  &  \bfseries{black} & very unstable rank \\
    \hline
    \end{tabular}
  \end{center} 
  \caption{Colour code of the RSI percentage text following the first condition satisfied.}
  \label{tab:RSI}
\end{table}

In the IBS diagnosed patient, beyond the differences in dominance for the particular taxa, we still observe that the most dominant are the most rank stable. However, as opposed to the healthy individual results, far from presenting a \emph{rank stability island}, the medium-ranked taxa are very rank unstable, mostly due to transient (often one or two consecutive samples) but deep drops in their relative abundance, which are usually happening more than twice along the time series. That is, for instance, the case of \emph{Sphingobacteriales} with two non-consecutive samples dropping to 111th rank position. In other cases, the high rank instability comes from a rank fluctuation over all the time series, as for \emph{Streptococcaceae} and \emph{Burkholderiales}, which are ranking 26th and 29th respectively in the overall dominance axis but show very low RSI, and thus colored in black attending to Table \ref{tab:RSI}.

>>>>

%%%
\subsection*{Rank stability of the taxa}

\def \RSItable {
\scriptsize
	\begin{tabular}{cccr}
	\multicolumn{4}{c}{Colour code for the RSI percentage column} \\
    \hline
    Case  &  Condition  &  Colour  &  Description \\
    \hline
    1  &  $1\ge{\rm RSI}>0.99$  & \textcolor{blue}{blue} & constant rank \\ 
    2  &  $0.99\ge{\rm RSI}>0.90$  &  \textcolor{green}{green}  & highly stable rank \\
    3  &  $0.90\ge{\rm RSI}>0.75$  &  \textcolor{orange}{orange} & moderately stable rank \\
    4  &  $0.75\ge{\rm RSI}>0.25$  &  \textcolor{red}{red} & unstable rank \\
    5  &  $0.25\ge{\rm RSI}\ge0$  &  \bfseries{black} & very unstable rank \\
    \hline
    \end{tabular}	
}

The rank dynamics and stability plots in Figure \ref{fig:corrank_IBS_A} and \ref{fig:corrank_IBS_P2} show the variation in the rank with time for the most dominant taxa and their calculated Rank Stability Index (RSI, as discussed in Material and Methods) for the taxa of a healthy subject (individual \emph{A}, Figure \ref{fig:corrank_IBS_A}) and from a subject diagnosed with IBS (patient \emph{P2}, Figure \ref{fig:corrank_IBS_P2}) of the IBS study\cite{IBS}. The taxa are listed ordered by the accumulated frequency along the time series, so y-axis is an overall dominance axis for each sample set. Generally speaking, we observe that the most dominant taxa are the most rank stable.

\begin{figure}
	\centering
	%\vspace*{-10mm} % Corrects overbox of the figures
	\includegraphics[width=0.99\textwidth]{results/corrank/IBS_h_A_amplicons_family_Rank.pdf}
	\RSItable	
	\caption{Rank variation throughout time for the most dominant elements (taxa) and their calculated Rank Stability Index (as shown in Material and Methods) for samples from a healthy subject studied in our lab\cite{IBS}.}
	\label{fig:corrank_IBS_A}
\end{figure}

\begin{figure}
	\centering
	%\vspace*{-10mm} % Corrects overbox of the figures
	\hspace*{-4.5mm}\includegraphics[width=0.99\textwidth]{results/corrank/IBS_P2_Metatranscriptores_Rank.pdf}
	\RSItable
	\caption{Rank variation throughout time for the most dominant elements (taxa) and their calculated Rank Stability Index for samples from a subject diagnosed with irritable bowel syndrome studied in our lab\cite{IBS}.}
	\label{fig:corrank_IBS_P2}
\end{figure}

Nevertheless, in the particular case of the healthy individual in Figure \ref{fig:corrank_IBS_A}, \emph{Burkholderiales} and \emph{Betaproteobacteria} (taxa ordered as 18th and 25th in the dominance axis) show comparatively very low rank stability regarding similar dominant taxa while, on the other hand, \emph{Comamonadaceae}, \emph{Lactobacillaceae}, \emph{Fusobacteriaceae}, \emph{Aerococcaceae} and \emph{Carnobacteriaceae} show higher stability than other more dominant taxa, forming a kind of \emph{rank stability island} for medium-ranked taxa around position 40 in the dominance axis, and thus colored in orange, following the color criteria shown in the table included in Figure \ref{fig:corrank_IBS_A}, since they show a moderately stable RSI.

In the IBS diagnosed patient of Figure \ref{fig:corrank_IBS_P2}, beyond the differences in dominance for the particular taxa, we still observe that the most dominant are the most rank stable. However, as opposed to the healthy individual results, far from presenting a \emph{rank stability island}, the medium-ranked taxa are very rank unstable, mostly due to transient (often one or two consecutive samples) but deep drops in their relative abundance, which are usually happening more than twice along the time series. That is, for instance, the case of \emph{Sphingobacteriales} with two non-consecutive samples dropping to 111th rank position. In other cases, the high rank instability comes from a rank fluctuation over all the time series, as for \emph{Streptococcaceae} and \emph{Burkholderiales}, which are ranking 26th and 29th respectively in the overall dominance axis but show very low RSI, and thus colored in black attending to the color criteria shown in the table included in Figure \ref{fig:corrank_IBS_P2}.
